First steps towards a formalization of Forcing

### 2018-08-17 LSFA/names.tex
Explicación para wfrec y Hv

\subsection{Names}
\label{sec:names}
We formalize the function $\val$ that allows to
construct the elements of the generic extension $M[G]$ from elements
of the ctm $M$ and the generic filter $G$. The definition of $\val$
can be written succintly as
%
\begin{equation}\label{eq:def-val}
\val(G,\tau)\defi \{\val(G,\sigma) : \exists p\in \PP.\,
(\lb\sigma,p\rb\in\tau \y p\in G)\}.
\end{equation}
%
As with any recursive definition, in order to justify that the
definition makes sense (i.e., that it defines anything at all) one has
to make explicit the well-founded relation which is used. In this
case, we use the relation $\mathit{ed}$:
\[
x \mathrel{\mathit{ed}} y \iff \exists p . \lb x,p\rb\in y.
\]
Recall that in $\ZFC$, an ordered pair $\lb x,y \rb$ is the set
$\{\{x\},\{x,y\}\}$. It is trivial to deduce the well-foundedness of
$\mathit{ed}$ from the fact that $\in$ is well-founded, which follows
from the Foundation Axiom.

The formalization of $\val$ in Isabelle/ZF involves the function
\isa{wfrec},\footnote{Notice that this form of recursive
  definitions is more general than the one used in the previous
  section to define \isa{dc{\isacharunderscore}witness}.} which takes as an argument a
well-founded relation of type \isa{i}, i.e.\ a set of pairs. Therefore
we restrict $\mathit{ed}$ above to a set.
%
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ edrel\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}{\isacharless}x{\isacharcomma}y{\isachargreater}\ {\isasymin}\ A{\isacharasterisk}A\ {\isachardot}\ x\ {\isasymin}\ domain{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}
%
Since \isa{edrel(A)} is a subset of a  well-founded relation (the
transitive closure of the membership relation restricted to \isa{A}),
then it is well-founded as well.

@@

\noindent We now turn to the definition of $\val$. In order to use
\isa{wfrec} we need to define a functional \isa{Hv}.

\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ Hv\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}Hv{\isacharparenleft}G{\isacharcomma}x{\isacharcomma}f{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ f{\isacharbackquote}y\ {\isachardot}{\isachardot}\ y{\isasymin}\ domain{\isacharparenleft}x{\isacharparenright}{\isacharcomma}\ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}y{\isacharcomma}p{\isachargreater}\ {\isasymin}\ x\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
\isanewline
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ val\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}edrel{\isacharparenleft}eclose{\isacharparenleft}M{\isacharparenright}{\isacharparenright}{\isacharcomma}\ {\isasymtau}{\isacharcomma}\ Hv{\isacharparenleft}G{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
Then we can recover the recursive expression~(\ref{eq:def-val}) thanks to the
following lemma:
%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ def{\isacharunderscore}val{\isacharcolon}
\isanewline
\ \ \ \ \ \ \ \  {\isachardoublequoteopen}x{\isasymin}M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}\ {\isacharcomma}\ {\isasymexists}p{\isasymin}P\ {\isachardot}\ \ {\isasymlangle}t{\isacharcomma}\ p{\isasymrangle}{\isasymin}x\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}

>>>>

\subsection{Names}
\label{sec:names}
We formalize the function $\val$ that allows to
construct the elements of the generic extension $M[G]$ from elements
of the ctm $M$ and the generic filter $G$. The definition of $\val$
can be written succintly as a recursive equation
%
\begin{equation}\label{eq:def-val}
\val(G,\tau)\defi \{\val(G,\sigma) : \exists p\in \PP.\,
(\lb\sigma,p\rb\in\tau \y p\in G)\}.
\end{equation}
%
The justification that $\val$ is well-defined comes from a general
result (transfinite recursion on well-founded
relations~\cite[p. 48]{kunen2011set}). Given a well-founded relation
$R \subseteq A \times A$ and a functional
$H : A \times (A \to A) \to A$, the principle asserts the existence of
a function $F : A \to A$ satisfying
$F(a) = H(a,F\uparrow\,(R^{-1}(a)))$. This principle is formalized in
Isabelle/ZF and one can use the operator \isa{wfrec}\footnote{Notice
  that this form of recursive definitions is more general than the one
  used in the previous section to define
  \isa{dc{\isacharunderscore}witness}.} to define functions using
transfinite recursion. To be precise, \isa{wfrec :: [i, i, [i,i]=>i]
  => i} is a slight variation, where the first argument is the
relation, the third is the functional, and the second corresponds to
the argument of $F$. Notice that the relation and the function argument
of the functional are internalized as terms of type \isa{i}.

In our case the functional $H$ takes an additional argument for the
parameter $G$:
\[
  H(G,y,f) = \{f(x) : x \in dom(y) \y \exists p\in \PP.\,
(\lb x,p\rb\in y \y p\in G)\}
\]
while the relation is given by:
\[
x \mathrel{\mathit{ed}} y \iff \exists p . \lb x,p\rb\in y.
\]
Recall that in $\ZFC$, an ordered pair $\lb x,y \rb$ is the set
$\{\{x\},\{x,y\}\}$. It is trivial to deduce the well-foundedness of
$\mathit{ed}$ from the fact that $\in$ is well-founded, which follows
from the Foundation Axiom.

In the formalization of $\mathit{ed}$ is restricted to a set,
so to have a term of type \isa{i}:
% 
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ edrel\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}{\isacharless}x{\isacharcomma}y{\isachargreater}\ {\isasymin}\ A{\isacharasterisk}A\ {\isachardot}\ x\ {\isasymin}\ domain{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}
%
Since \isa{edrel(A)} is a subset of a  well-founded relation (the
transitive closure of the membership relation restricted to \isa{A}),
then it is well-founded as well.

@@

\noindent The formalizaton of the functional $H$ is straightforward and $\val$ is defined using \isa{wfrec}:
%
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ Hv\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}Hv{\isacharparenleft}G{\isacharcomma}y{\isacharcomma}f{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ f{\isacharbackquote}x\ {\isachardot}{\isachardot}\ x{\isasymin}\ domain{\isacharparenleft}y{\isacharparenright}{\isacharcomma}\ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}x{\isacharcomma}p{\isachargreater}\ {\isasymin}\ y\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
\isanewline
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ val\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}edrel{\isacharparenleft}eclose{\isacharparenleft}M{\isacharparenright}{\isacharparenright}{\isacharcomma}\ {\isasymtau}{\isacharcomma}\ Hv{\isacharparenleft}G{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
Then we can recover the recursive expression~(\ref{eq:def-val}) thanks to the
following lemma:
%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ def{\isacharunderscore}val{\isacharcolon}
\isanewline
\ \ \ \ \ \ \ \  {\isachardoublequoteopen}x{\isasymin}M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}\ {\isacharcomma}\ {\isasymexists}p{\isasymin}P\ {\isachardot}\ \ {\isasymlangle}t{\isacharcomma}\ p{\isasymrangle}{\isasymin}x\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}
